---
title: 'Class 2: Ways to Access Data in R'
author: "Jenine Harris"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4) 
library(package = "tidyverse")
```

## Options for accessing data in R 

* Use a built-in data source 
* Type in data 
* Import a local data set saved on your computer 
* Import online data  

## Use a built-in data source 

* There are several data sources that come with R and are accessible every time you open R. 
* There is no need to **open** or **load** these data sources, they are just ready and waiting for you. 
* To get a list of all the built-in data sources, use the R function `data()`. 
* One data source is `USArrests` which has the numer of murders, assaults, rapes, and the population size of each state in the US. 
* One way to get a glimpse of this data set, use the `head()` function.

```{r}
# show the first few cases of USArrests data
head(x = USArrests)
```

## Type in your own data 

* While it is probably not ideal for most data sets, typing data directly into R is possible. 
* For example, here are the number of people in this webinar who teach in each software package. 

```{r}
 # enter the data 
software <- c("SPSS", "Excel", "SAS", "Stata", "R")
numPart <- c(107, 29, 13, 12, 2)

# combine into a data set
softwareJobs <- data.frame(software, numPart)
```

## Make a graph

Once the data are entered, use R functions like `ggplot()` to explore the data.  

```{r}
# make a graph with ggplot
softwareJobs %>% 
  ggplot(aes(x = reorder(software, numPart), 
             y = numPart)) +
  geom_col() +
  labs(x = "Software", 
       y = "Number of webinar participants using to teach") +
  theme_minimal() +
  coord_flip()
```

## Import a local data set saved on your computer 

* R has import functions for most file types used to save data. 
* For example, `read.spss()` can import SPSS files while SAS files can be opened with `read_sas()`. 
* The data imported here are in a **.csv** file.
* These data are from the The Foundation for AIDS Research (amfAR) website and are the subject of Chapter 9 in *Statistics with R: Solving Problems Using Real World Data*. 

```{r}
# bring in syringe program data 
syringeProgDist <- read.csv("dist_ssp_amfar_ch9.csv")
```

## Explore the data 

```{r}
# summarize the data 
summary(object = syringeProgDist)
```

## Make a fancy graph

* The data are at the county-level and have variables for how far the nearest syringe program is and whether the county is a metropolitan (city) area or non-metropolitan area. 
* To show off a little of what R is capable of, here is one way to examine the distribution of these data by groups (thanks to Shelly Cooper). 

```{r eval = FALSE}
# fancy graph showing distance to nearest syringe program
# by type of county in the US 
syringeProgDist %>%
  ggplot(aes(x = metro, y = dist_SSP, fill = metro)) +
  geom_violin(aes(color = metro), fill = "white", alpha = .8) + 
  geom_boxplot(aes(fill = metro, color = metro), width = .2, alpha = .3) +
  geom_jitter(aes(color = metro), alpha = .4) +
  labs(x = "Type of county", 
       y = "Miles to syringe program") +
  scale_fill_manual(values = c("#78A678", "#7463AC"), guide = FALSE) + 
  scale_color_manual(values = c("#78A678", "#7463AC"), guide = FALSE) +
  theme_minimal() +
  coord_flip()
```

## See the fancy graph 

```{r echo = FALSE}
# fancy graph showing distance to nearest syringe program
# by type of county in the US 
syringeProgDist %>%
  ggplot(aes(x = metro, y = dist_SSP, fill = metro)) +
  geom_violin(aes(color = metro), fill = "white", alpha = .8) + 
  geom_boxplot(aes(fill = metro, color = metro), width = .2, alpha = .3) +
  geom_jitter(aes(color = metro), alpha = .4) +
  labs(x = "Type of county", 
       y = "Miles to syringe program") +
  scale_fill_manual(values = c("#78A678", "#7463AC"), guide = FALSE) + 
  scale_color_manual(values = c("#78A678", "#7463AC"), guide = FALSE) +
  theme_minimal() +
  coord_flip()
```

## Import a data set from the internet 

* Many government agencies, non-profits, foundations, universities, authors, and others post data online. 
* While there can be challenges bringing data in from online (e.g., what if the URL changes?), it also ensures that the data used are the most current data and the work is more likely to be reproducible. 
* The data set accessed here is used in Chapter 6 of *Statistics For People Who Think They Hate Statistics* and is about student engagement on days of the week in class.

```{r}
# bring in student engagement data from GitHub repository
daysWeek <- read.csv("https://raw.githubusercontent.com/jenineharris/R-you-ready-demo/master/ch6ds3.csv")
```

## Explore the data  

```{r}
# summarize the data source 
summary(object = daysWeek)
```

## Transform the data and make a fancy graph 

```{r}
# convert day of the week variables to one factor variable
# graph using density plots to compare days
daysWeek %>% 
  pivot_longer(cols = ends_with("day"), names_to = "day.of.week", 
               values_to = "engage.score") %>% 
  ggplot(aes(x = engage.score, fill = day.of.week)) +
  geom_density(alpha = .6) + theme_minimal() +
  labs(x = "Student Engagement Score") +
  scale_fill_discrete(name = "Day of Week")
```





